Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updated

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing the steps of \textit{ExpertInLoop} algorithm. The dashed edges show all the potential edges that \textit{Expand} algorithm could add in each step.
	}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textit{ExpertInLoop} algorithm. As automated algorithms only recover
		the CPDAG, we use the best and worst scoring orientation of the CPDAG 
		to get the range. The \textit{ExpertInLoop} shows values for various
		$ \alpha_{\textit{eff}} $ values corresponding to $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $.}
	\label{fig:shd_sid}
\end{figure}

@@

We analyzed the performance of \textit{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated data from known DAG
structures and use the algorithms to recover the original DAG. We start by
generating a random DAG on $10$ nodes and use linear models with random effects
to generate the data. We vary the edge probability of the DAG to generate DAGs
with varying density. We repeat the data generation $30$ times for each density
value for the DAG. We compare the learned DAG to the original DAG using two
metrics: Structural Hamming Distance (SHD) and Structural Intervention Distance
(SID) \citep{Peters2015}.

@@

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case scenarios.

To use the \textit{ExpertInLoop} algorithm, we need two oracles: a d-separation
oracle and an ancestral oracle. We use a conditional independence test (partial
correlation if using linear data) for determining the d-separation with a
p-value threshold of $ 0.05 $. For the ancestral oracle, we simulated the
strong ancestral oracle. We vary the accuracy of our oracle to test the
performance for varying oracle accuracies. Given an accuracy level $ \alpha $
of the oracle, the oracle returns:

\begin{equation}
	\begin{split}
		x &= \textnormal{rand}([0, 1]) \\
		O(\alpha) &= \begin{cases} 
			M \rightarrow Y, & \textnormal{if  } x <= \alpha \\
			\textnormal{rand}(M \rightarrow Y, M \leftarrow Y, None) & \textnormal{otherwise} \\
				\end{cases} \\
	\end{split}
\end{equation}

If the oracle returns None for any pair of variables, an edge between this pair
is not suggested to the oracle in future iterations. We also make sure that
only edges which do not form a cyclic in the graph are used to select the next
potential edge. There is a possibility that this greedy expert gets stuck at
incorrect graph structures as it does not do any backtracking to fix its
earlier mistakes. This procedure is outlined in the Algorithm 1.
\todo[inline]{Show an algorithm of how this expert method is working}

>>>>

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing each iteration of the \emph{ExpertInLoop} algorithm. The dashed edges show all the potential edges that \emph{Expand} step could add in each iteration.}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\emph{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\emph{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

@@

The \emph{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the orientation or
non-existence of edges between pairs of variables. For answering the
d-separation queries, we use a CI test, the partial correlation test with a
p-value threshold of $ 0.05 $. Secondly, to tell the direction between edges or
non-existence of edges, we simulate an expert with accuracy $ \alpha $ as
follows:

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		E(\alpha) &= \begin{cases} 
			\mathrm{TrueDir}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
	\end{split}
\end{equation}

where given the true DAG, $ G $:

\begin{equation}
	\begin{split}
	\mathrm{TrueDir}(X, Y) &= \begin{cases}
					X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G \\
					Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}

An important point to note here is that the effective accuracy of $ E $ is
higher than $ \alpha $ as even when giving a random answer, there is a $ 1 $ in
$ 3 $ chance that the answer is correct. The effective accuracy is:

We analyzed the performance of \emph{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated $ 500 $ samples from
known DAGs and use the algorithms to recover the original DAG. We start by
generating a random DAG on $10$ nodes and use linear models with random effects
to generate the data. We vary the edge probability of the DAG to generate DAGs
with varying density. We repeat the data generation $30$ times for each density
value for the DAG. We compare the learned DAG to the original DAG using two
metrics: Structural Hamming Distance (SHD) and Structural Intervention Distance
(SID) \citep{Peters2015}.

@@

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case
scenarios. The results are shown in Figure~\ref{shd_sid}. For SHD, the
performance of \emph{ExpertInLoop} is comparable to the automated algorithms
for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, and performs better for
higher $ \alpha $ values. For SID, the \emph{ExpertInLoop} performs better in
denser DAGs.
